Patterns of floral and chromosome evolution in Onagraceae: a fossil-calibrated supermatrix analysis

### 2014-05-15 manuscript.tex
rough draft intro complete

Introduce supermatrix analyses....

Onagraceae, the evening primroses, are a well-studied family
of flowering plants with 22 described genera and around 660 species \citep{wagner2007revised}.
The family is found on all continents except Antartica,
and pollen fossils date Onagraceae to the Upper Cretaceous \citep{grimsson}.
Onagraceae is well placed in the order Myrtales,
and previous molecular phylogenetic studies find it most closely related to Lythraceae \citep{conti1997, Sytsma2004}.
Though there have been many genus-level molecular phylogenetic studies in Onagraceae
\citep{Berry2004, Evans2009, Hoggard2004, Xie2009, Baum1994, Wagner2005}, 
the most recent family-level analyses \citep{Levin2003, Levin2004}
used only one or two exemplars from some genera. 
Furthermore, the historical biogeography, divergence times of the major clades, and 
family-level patterns and processes of diversification remain poorly understood.
With the wealth of data available, Onagraceae is an
ideal group to apply a data-mining supermatrix approach
with the goal of constructing a densely sampled family-level phylogeny.

Here, I use a well-resolved supermatrix phylogeny of Onagraceaea to
answer qestions such as:
1) Are the taxonomic groups described in \citet{wagner2007revised} monophyletic?
2) When did the major clades diverge?
3) What is the pattern of chromosome number
evolution?
4) What are the patterns of petal color and petal number evolution, and are their 
evolution correlated?
Finally, I also discuss ways in which these analyses can be extended
to further examine the patterns and processes of
evolution in Onagraceae.

@@

discuss monophyly of tribe Epilobieae, which was supported in 
maximum likelihood analysis but not supported in Bayesian MCC, could be due to
fossil constraints.....

>>>>

In pursuit of large-scale evolutionary questions,
biologists are increasingly using massive phylogenetic
datasets to reconstruct ever-growing portions of the tree of life. 
A common approach to building large phylogenetic trees has been
the supermatrix approach, in which multiple datasets are
combined and analyzed simultaneously \citep{DeQueiroz2007}.
However, assembling and utilizing supermatrices is not trivial
due to challenges such as determining homology of molecular
sequences, assembling chimeric operational taxonomic units, 
limiting the amount of missing data,
and managing data integrity and computational feasibility \citep{Roure2007, Yan2005}.
Despite these difficulties, considerable bioinformatic advances 
have made supermatrix construction more common,
particularly in large-scale evolutionary plant studies.
For example, \citet{Smith2011}
inferred a phylogeny of 55473 seed plants,
\citet{Hinchliff2013} estimated the phylogeny of 1500 members of Cyperaceae,
and \citet{Sarkinen2013} built a time calibrated tree of Solanaceae with 1076 tips.

With over 5000 DNA sequences available in GenBank from previous studies, 
the plant family Onagraceae is an
ideal group to apply a data-mining supermatrix approach.
Onagraceae, the evening primrose family, is a well-studied 
monophyletic clade
of flowering plants with 22 described genera and around 660 species \citep{wagner2007revised}.
The family is found on all continents except Antartica,
and pollen fossils date Onagraceae to the Upper Cretaceous \citep{grimsson}.
Onagraceae falls within the order Myrtales,
and previous molecular phylogenetic studies find it most closely related to Lythraceae \citep{conti1997, Sytsma2004}.
Though there have been many genus-level molecular phylogenetic studies in Onagraceae
\citep{Berry2004, Evans2009, Hoggard2004, Xie2009, Baum1994, Wagner2005}, 
the most recent family-level analyses \citep{Levin2003, Levin2004}
used only one or two exemplars from some genera. 
Furthermore, the historical biogeography, divergence times of the major clades, and 
family-level patterns and processes of diversification remain poorly understood
and untested over a molecular phylogeny.

Here, I use a well-resolved and densely-sampled supermatrix phylogeny of Onagraceae to
test whether the taxonomic groups described in \citet{wagner2007revised} are monophyletic,
and to answer the following questions:
1) When did the major clades diverge?
2) What is the pattern of chromosome number evolution?
3) What are the patterns of petal color and petal number evolution, and are their 
evolution correlated?
Lastly, I consider ways in which these analyses can be extended
to further examine the evolution of Onagraceae, 
and discuss the novel method used here to construct the supermatrix.

@@

discuss monophyly of tribe Epilobieae, which was supported in 
maximum likelihood analysis but not supported in Bayesian MCC, could be due to
fossil constraints on genus Epilobium, it may be that the fossil described should
have constrained Epilobium + Chamerion due to taxonomic changes.....
need to check fossil synapomorphies to see where it should be placed!

Petal color and petal number were found to follow a pattern 
of correlated evolution (p = 0.00), 
suggesting concerted shifts in floral traits 
may play an important role in shaping diversification of Onagraceae. 
Further analyses should be performed to test 
whether shifts in floral traits are correlated with pollinator shifts, 
and to determine how these shifts affect diversification rates. 
Furthermore, this work should be extended to test historical 
biogeographical hypotheses that explain the intercontinental disjunct distribution of Onagraceae.

For example, the PhyLoTA Browser 
provides a web interface to view all GenBank sequences within
taxonomic groups clustered into homologs \citep{Sanderson2008}.
A different approach is implemented in the program PHLAWD,
which constructs clusters of homologs around
guide sequences provided by the user \citep{Smith2009}.
However, the method used by SUMAC combines both approaches;
the user can perform an exploratory clustering of all GenBank sequences
within a taxonomic group, and also provide guide sequences to build
homologous clusters in a more targeted manner. SUMAC provides
a framework with which GenBank can be repeatedly mined 
using different taxonomic groups and/or different guide sequences
and the resulting data matrices can be compared.

describe some recently developed metrics that can
be used to assess the informativeness of phylogenetic supermatrices.
